{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-19 thesis/education.tex
write quicksand design, beginnings of proof stuff
daves final revisions to educ

Since the Spring 2015 semester I have given a guest lecture in 15-410
to recruit students to participate in the user study.
The 50-minute lecture is given 1 week into the 2.5-week-long P2 project,
approximately when the students should be getting child threads running in {\tt thr\_create()}
and experiencing concurrency bugs for the first time.
It introduces the research subject abstractly
using an example ``Paradise Lost'' bug from a previous lecture \cite{paradise-lost},
explains how Landslide works concretely,
shows a short demo of effortlessly using Landslide to find the example bug,
and provides the necessary IRB legalese about the risks and benefits of participation.
The most recent lecture slides are available on the course website at
\url{http://www.cs.cmu.edu/~410-s18/lectures/L14_Landslide.pdf},
and all semesters' editions at
\url{https://github.com/bblum/talks/tree/master/landslide-lecture}.

@@

At both schools students then signed up using a Google form I emailed them,
which upon completion linked them to the Landslide user guide,
which is available at
\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-p2.pdf}
(CMU version)
and
\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-psu.pdf}
(PSU version)
and
\url{https://github.com/bblum/talks/tree/master/irb}
(both versions).

@@

The following tests were released to CMU students:
\begin{itemize}
	\item {\tt broadcast\_test}:
		Tests the {\tt cond\_broadcast()} signalling path with a single waiter.
	\item {\tt mutex\_test}:
		Tests student mutexes under 2 threads with 2 iterations
		(the \revisionminor{second} iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
		This test uses the {\tt TESTING\_MUTEXES}
		described in \cref{sec:landslide-staticconfig}
		to enable data-race preemption points within the mutex implementation.
	\item {\tt paradise\_lost}:
		Written for the sake of the Landslide lecture demo
		(\cref{sec:education-pebbles-recruiting}).
		Tests for the Paradise Lost bug by attempting to break mutual exclusion.
	\item {\tt paraguay}:
		Copied directly from the P2 test suite;
		tests for proper handling of seemingly ``spurious'' wakeups in {\tt cond\_wait()}.
		Written by Michael J. Sullivan.
	\item {\tt rwlock\_downgrade\_read\_test}:
		Copied directly from the P2 test suite;
		tests for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()}
		\revisionminor{(a function which atomically converts a lock held in exclusive write mode to shared read mode)}.
		Written by me (as a TA).
	\item {\tt thr\_exit\_join}:
		Copied directly from the P2 test suite;
		tests for a variety of problems between {\tt thr\_exit()} and {\tt thr\_join()},
		but especially for memory issues pertaining to stack deallocation.
\end{itemize}
\vspace{1em}

@@

The (unpatched versions of) the first four tests are available at
\url{https://github.com/bblum/pintos} (a fork of the main Pintos basecode repository).
%\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
The fifth test is available at
\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.

@@

\section{Evaluation}

@@

\begin{itemize}
	\item How many bugs, and of what severities, does Landslide help students find and fix before submitting their code?
	\item Does Landslide use result in higher quality submissions, whether directly for P2, or for subsequent projects as well?
	\item Do students feel the experience is worthwhile, as compared to stress testing?
	\item How well does Landslide apply to operating systems projects outside of CMU?
\end{itemize}
\vspace{1em}

@@

{\bf Statistical significance.}
\Cref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these 4 comparsons \revisionminor{as} I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons,
\revisionminor{which do not address that alternative hypothesis,
were significant,
while the cross-semester comparisons,
which attempted to do so,
failed to confidently reject even the null hypothesis.}
However, the distribution difference being smaller in P3 than in P2
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

% TODO check figure placement
\begin{table}[h]
	\begin{center}
		\small
		% note, smth like fisher's method which "combines independent studies into a meta p value"
		% cannot be applied for 2 reasons - 1, these are not independent since the same studence were
		% running landslide on diff test cases; 2, the hypotheses tested are different for each test
	\begin{tabular}{r||cc|cc|c|cc}
		& \multicolumn{2}{c|}{\bf Landslide users} & \multicolumn{2}{c|}{\bf Non-users} & & \multicolumn{2}{c}{\bf Significance} \\
		\bf Bug name & \bf \#correct & \bf \#buggy & \bf \#correct & \bf \#buggy & \bf $\Delta$\%correct & $p$ & \bf cutoff \\
		\hline
		% calculated in R with (rest done similarly)
		% fisher.test(rbind(c(11,10),c(0,16)))$p.value --> 0.0006143355
		Exit UAF  & 11	& 10	& 0	& 16	& $+52.4\%$	& 0.00061	& 0.0125 \\
		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
		% Can't name the group here, but if you're me and have the analysis file,
		% it's the one from f17 with 3 mutex_test logs in their "u" snapshot.
		% It seems like a mysterious older version of mutex.c was present between their "a" version
		% (which gets fully verified very quickly btw) and their "u" version (for which landslide finds
		% the bug in just a few minutes, no way they could be running the same version; plus the reported
		% data races and so on in the id-log don't line up to the "u" mutex.c source code).
		% That version was correct, then they updated it (introducing the bug) and ran a bunch of other tests,
		% but didn't rerun mutex_test to check their update.
		Mutex     & 17	& 1	& 19	& 0	& $-05.6\%$	& 0.49 & 0.0125 \\
		Paraguay  & 20	& 1	& 14	& 2	& $+07.7\%$	& 0.57	& 0.0125 \\
		% nb this is any downgrade related ding not just dgr tested ones
		% nb 1 is msising bc didn't impl rwlock at all
		% nice p value
		Downgrade & 10	& 3	& 19	& 4	& $-05.7\%$	& 0.69	& 0.0125 \\ % TODO verify these buggers by eye
		\hline
		\bf Total & 58	& 15	& 52	& 22	& $+09.2\%$	& 0.26 & 0.05 \\
	\end{tabular}
	\end{center}
	\caption[Correlation of student Landslide use with solving certain bugs.]
		{
		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
		%$\Delta P$ stands for the increased probability that a student submits a correct P2
		%(with respect to each bug)
		%given that they use Landslide, versus not using Landslide; i.e.,
		%$P(\mathsf{correct}|\mathsf{landslide})-
		%P(\mathsf{correct}|\mathsf{non\text{-}user})$.
		Note that the totals in the bottom row double-count students,
		which makes sense only if you believe the incidences of each bug in a given submission are independent.
		}
	\label{tab:eval-common-bugs}
\end{table}

@@

\subsubsection{Reasons worthwhile}

@@

\begin{itemize}
	\item A lecture, polished over seven semesters,
		for introducing stateless model checking as a concurrency testing approach
		in high-level undergraduate classes
		(with many slides admittedly specific to Landslide as they may be)
		(\cref{sec:education-pebbles-recruiting}).
	\item A system for automatically instrumenting student thread library (CMU, PSU) and kernel (U. Chicago)
		project implementations to be compatible with stateless model checking
		(\cref{sec:education-pebbles-instrumentation},
		\cref{sec:education-pintos-instrumentation}).
	\item An experimental procedure for evaluating debugging and verification tools
		as student-accessible teaching aids,
		including a survey for measuring the more intangible human experience
		(\cref{sec:education-survey-pebbles},
		\cref{sec:education-survey-pintos}).
	\item A large evaluation spanning nine semesters of operating systems classes
		(\cref{sec:education-eval-bugfinding}),
		two of which were held at other universities besides CMU,
		in which 145 students used Landslide,
		a majority of which found bugs (\cref{sec:education-eval-bugfinding}),
		and a vast majority of which expressed via the survey that they had a good experience
		(\cref{sec:education-eval-survey}).
	\item A thorough statistical analysis of Landslide's effect on student submission quality
		(\cref{sec:education-eval-grades}),
		evaluating both impact on ultimate project grades
		and student tendency to submit any of four well-known case study bugs.
		Ultimately, while positive trends were observed,
		this study concluded that a statistically-significant conclusion could not be drawn
		while simultaneously compensating for several obvious sources of possible bias.
		Nevertheless, this level of statistical rigor is not yet widespread in computer science research,
		so I would hope this study may set a precedent that help mature the field.
	\item Concrete action taken to establish Landslide
		as a permanent fixture of 15-410's grading process at CMU,
		student-oriented user guides should the course staff wish it to remain available as a debugging tool,
		as well as general recommendations for any Pintos-based OS class
		at other universities to use it for grading as well
		(\cref{sec:education-future-use}).
\end{itemize}
}

>>>>

%Since the Spring 2015 semester
From the Spring 2015 semester through Spring 2018,
I have given a guest lecture in 15-410
to recruit students to participate in the user study.
The 50-minute lecture is given 1 week into the 2.5-week-long P2 project,
approximately when the students should be getting child threads running in {\tt thr\_create()}
and experiencing concurrency bugs for the first time.
It introduces the research subject abstractly
using an example ``Paradise Lost'' bug from a previous lecture \cite{paradise-lost},
explains how Landslide works concretely,
shows a short demo of effortlessly using Landslide to find the example bug,
and provides the necessary IRB legalese about the risks and benefits of participation.
The most recent lecture slides are available on the course website at
\url{https://www.cs.cmu.edu/~410-s18/lectures/L14_Landslide.pdf},
and all semesters' editions at
\url{https://github.com/bblum/talks/tree/master/landslide-lecture}.

@@

At both schools students then signed up using a Google form I emailed them,
which upon completion linked them to the Landslide user guide,
which is available at
\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-p2.pdf}
(CMU version)
and at % these "at"s can optionally be adjusted to make url line breaks better
\url{http://www.contrib.andrew.cmu.edu/~bblum/landslide-guide-psu.pdf}
(PSU version)
and at
\url{https://github.com/bblum/talks/tree/master/irb}
(both versions).

@@

The following tests were released to CMU students:
\begin{itemize}
	\item {\tt broadcast\_test}:
		Tests the {\tt cond\_broadcast()} signalling path with a single waiter.
	\item {\tt mutex\_test}:
		Tests student mutexes under 2 threads with 2 iterations
		(the \revisionminor{second} iteration serves to expose problems with {\tt mutex\_unlock()} as well as {\tt mutex\_lock()}).
		This test uses the {\tt TESTING\_MUTEXES}
		described in \cref{sec:landslide-staticconfig}
		to enable data-race preemption points within the mutex implementation.
	\item {\tt paradise\_lost}:
		Written for the sake of the Landslide lecture demo
		(\cref{sec:education-pebbles-recruiting}).
		Tests for the Paradise Lost bug by attempting to break mutual exclusion.
	\item {\tt paraguay}:
		Copied directly from the P2 test suite;
		tests for proper handling of seemingly ``spurious'' wakeups in {\tt cond\_wait()}.
		Written by Michael J. Sullivan.
	\item {\tt rwlock\_downgrade\_read\_test}:
		Copied directly from the P2 test suite;
		tests R/W locks for mutually-exclusive and deadlock-free {\tt rwlock\_downgrade()},
		\revisionminor{which should atomically convert a lock held in exclusive write mode to shared read mode}.
		Written by me (during my time as a TA).
	\item {\tt thr\_exit\_join}:
		Copied directly from the P2 test suite;
		tests for a variety of problems between {\tt thr\_exit()} and {\tt thr\_join()},
		but especially for memory issues pertaining to stack deallocation.
\end{itemize}
\vspace{1em}

@@

The (unpatched versions of) the first four tests are available at
\url{https://github.com/bblum/pintos} (a fork of the main Pintos basecode repository).
%\url{https://github.com/Berkeley-CS162/group0/tree/master/pintos/src/tests}.
The fifth test is available in the Landslide repository at {\tt pebsim/pintos/priority-donate-multiple.c}.
%\url{https://github.com/bblum/landslide/blob/master/pebsim/pintos/priority-donate-multiple.c}.

@@

\section{Evaluation}
\label{sec:education-eval}

@@

\vspace{0.5em}
\begin{enumerate}
	\item How many bugs, and of what severities, does Landslide help students find and fix before submitting their code?
	\item Does Landslide use result in higher quality submissions, whether directly for P2, or for subsequent projects as well?
	\item Do students feel the experience is worthwhile, as compared to stress testing?
	\item How well does Landslide apply to operating systems projects outside of CMU?
\end{enumerate}
\vspace{0.5em}

@@

{\bf Statistical significance.}
\Cref{tab:table-of-ze-studence} presents more detailed statistics from these six distributions.
To assess whether the differences are significant,
% discrete == I used "version 2" of the AD metric that R outputs.
I use the $k$-sample discrete Anderson-Darling test \cite{anderson1952,anderson-darling}
provided by the R statistical programming environment \cite{r-lang,r-ksamples},
comparing each control group to its respective Landslide users group.
Anderson-Darling uses a weighted sum-of-squares distance metric to compare two empirical distribution functions (EDFs),
testing if their samples are likely to arise from the same underlying, though unspecified, distribution.%
\footnote{
I choose Anderson-Darling over the simpler Kolmogorov-Smirnov,
which computes only the maximum instantaneous distance between EDFs,
for its better sensitivity both to repeated deviations and to tail differences \cite{beware-kolmogorov-smirnov}.
}
I deem the difference in grade distributions significant when $p < 0.05$%
% FIXME: is this legit?
% should you be using the combined/grouped AD test instead?
\footnote{I choose not to correct for multiplicity among these four comparsons \revisionminor{as} I do in the next section
because each tests a different hypothesis.
},
although as in all $p$-value calculations,
this rejects only the null hypothesis,
not alternative hypotheses,
such as ``more skilled students were more likely to sign up to begin with''.
%
Ultimately, only the same-semester comparisons,
\revisionminor{which do not address that alternative hypothesis,
were significant,
while the cross-semester comparisons,
which attempted to do so,
failed to confidently reject even the null hypothesis.}
However, the distribution difference being smaller in P3 than in P2
further suggests, albeit informally,
that the larger impact in P2 is not attributable {\em only} to selection bias,
or else a similar difference should have been observed in the P3 distributions.
Another possible explanation could simply be that the P3 grading critera result in less grade variance overall,
but this pattern is at least consistent with the ``Landslide helps students submit better P2s'' hypothesis.

@@

% TODO check figure placement
\begin{table}[h]
	\begin{center}
		\small
		% note, smth like fisher's method which "combines independent studies into a meta p value"
		% cannot be applied for 2 reasons - 1, these are not independent since the same studence were
		% running landslide on diff test cases; 2, the hypotheses tested are different for each test
	\begin{tabular}{r||cc|cc|c|cc}
		& \multicolumn{2}{c|}{\bf Landslide users} & \multicolumn{2}{c|}{\bf Non-users} & & \multicolumn{2}{c}{\bf Significance} \\
		\bf Bug name & \bf \#correct & \bf \#buggy & \bf \#correct & \bf \#buggy & \bf $\Delta$\%correct & $p$ & \bf cutoff \\
		\hline
		% calculated in R with (rest done similarly)
		% fisher.test(rbind(c(11,10),c(0,16)))$p.value --> 0.0006143355
		Exit UAF  & 11	& 10	& 0	& 16	& $+52.4$\%	& 0.00061	& 0.0125 \\
		% I spent more hours than I should have looking at this landslide-user-submitted-bug group.
		% Can't name the group here, but if you're me and have the analysis file,
		% it's the one from f17 with 3 mutex_test logs in their "u" snapshot.
		% It seems like a mysterious older version of mutex.c was present between their "a" version
		% (which gets fully verified very quickly btw) and their "u" version (for which landslide finds
		% the bug in just a few minutes, no way they could be running the same version; plus the reported
		% data races and so on in the id-log don't line up to the "u" mutex.c source code).
		% That version was correct, then they updated it (introducing the bug) and ran a bunch of other tests,
		% but didn't rerun mutex_test to check their update.
		Mutex     & 17	& 1	& 19	& 0	& $-05.6$\%	& 0.49 & 0.0125 \\
		Paraguay  & 20	& 1	& 14	& 2	& $+07.7$\%	& 0.57	& 0.0125 \\
		% nb this is any downgrade related ding not just dgr tested ones
		% nb 1 is msising bc didn't impl rwlock at all
		% nice p value
		Downgrade & 10	& 3	& 19	& 4	& $-05.7$\%	& 0.69	& 0.0125 \\ % TODO verify these buggers by eye
		\hline
		\bf Total & 58	& 15	& 52	& 22	& $+09.2$\%	& 0.26 & 0.05 \\
	\end{tabular}
	\end{center}
	\caption[Correlation of student Landslide use with solving certain bugs.]
		{
		Correlation of student Landslide use with solving certain bugs in their final submission during Fall 2017 and Spring 2018 semesters.
		%$\Delta P$ stands for the increased probability that a student submits a correct P2
		%(with respect to each bug)
		%given that they use Landslide, versus not using Landslide; i.e.,
		%$P(\mathsf{correct}|\mathsf{landslide})-
		%P(\mathsf{correct}|\mathsf{non\text{-}user})$.
		Note that the totals in the bottom row double-count students,
		which makes sense only if you believe the incidences of each bug in a given submission are independent.
		}
	\label{tab:eval-common-bugs}
\end{table}

@@

\subsubsection{Reasons worthwhile}
\label{sec:education-reasons-worthwhile}

@@

\begin{itemize}
	\item A lecture, polished over seven semesters,
		for introducing stateless model checking as a concurrency testing approach
		in high-level undergraduate classes
		(with many slides admittedly specific to Landslide)
		(\cref{sec:education-pebbles-recruiting}).
	\item A system for automatically instrumenting student thread library (CMU, PSU) and kernel (U. Chicago)
		project implementations to be compatible with stateless model checking
		(\cref{sec:education-pebbles-instrumentation},
		\cref{sec:education-pintos-instrumentation}).
	\item An experimental procedure for evaluating debugging and verification tools
		as student-accessible teaching aids,
		including a survey for measuring the more intangible human experience
		(\cref{sec:education-survey-pebbles},
		\cref{sec:education-survey-pintos}).
	\item A large evaluation spanning nine semesters of operating systems classes
		(\cref{sec:education-eval-bugfinding}),
		two of which were held at other universities besides CMU,
		in which 145 students used Landslide,
		a majority of which found bugs (\cref{sec:education-eval-bugfinding}),
		and a vast majority of which expressed via the survey that they had a good experience
		(\cref{sec:education-eval-survey}).
	\item A thorough statistical analysis of Landslide's effect on student submission quality
		(\cref{sec:education-eval-grades}),
		evaluating both impact on ultimate project grades
		and student tendency to submit any of four well-known case study bugs.
		Ultimately, while positive trends were observed,
		this study concluded that a statistically-significant conclusion could not be drawn
		while simultaneously compensating for several obvious sources of possible bias.
		Nevertheless, this level of statistical rigor is not yet widespread in computer science research,
		so I would hope this study may set a precedent that help mature the field.
	\item Concrete action taken to establish Landslide
		as a permanent fixture of 15-410's grading process at CMU,
		student-oriented user guides should the course staff wish it to remain available as a debugging tool,
		as well as general recommendations for any Pintos-based OS class
		at other universities to use it for grading as well
		(\cref{sec:education-future-use}).
\end{itemize}
}
